\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-03 NAR_DRAFT.tex
working on discussion

\batcave is a post-call variant evalutation tool that uses MuTect output and the tumor-specific mutation profile to sharpen the posterior probability assigned to SNVs.
We demonstrated here that \batcave improves classification accuracy in simulated data, with the magnitude of the improvement dependent on the sequencing depth and the concentration of the mutation profile.
We report only precision-recall comparisons for real tumor data because in both cases the "gold" standard data sets consist of variants that were "called" by one or more variant detection pipelines, and confirmed to be present or absent.
These mutations are in some sense \textit{easy} to call, since at least one variant caller did in fact call them.
At the same time, there is no way to compute true and false negative fractions for these data sets.
As such, it is not surprising that the performance of MuTect and \batcave are similar because the universe of variants under consideration in these experiments are by definition "callable" by at least one variant detection pipeline, and MuTect is among the most sensitive and specific variant callers \cite{Griffith2015,Cibulskis2013}.
Deep sequencing experiments where a large random sample of uncalled potential variants, as well as a random sample of variant with no reads at all present, would give much-needed insight into the differences in statistical models in variant calling.

The \batcave algorithm focuses on the tri-nucleotide context and substitution type. 
There are many other types of \textit{genomic context} which can effect local mutation rates \cite{Buisson2019}.
These could be very useful in computing a local or regional mutation rate, as opposed to the genome-wide but tumor-specific mutation rate used by \batcave.

@@

\batcave approaches the set of variant calls returned from MuTect as two distinct sets of data.
The first is the set of mutations that pass all heuristics filters, and  have a very large posterior probability of being present given the observed read data.
These are high-probability variants generated from a distribution, the mutation profile, and measured with very small error.
This data allows us to reliably estimate the distribution of substitution types.
Variants with low probability given the read data are noisy measurements of samples from the same data generating distribution.
We use the estimated distribution of substitution types to adjust the prior probability of mutation for these lower probability variants.

@@

\begin{itemize}
  \item The fact that we can think of tumor variant calls as two sets of data, with different measurement error, derived from the same data generating process, makes this possible
  \item The mutation profile is estimated from the high-confidence (low measurement error) variants, and used as a prior for the low-confidence (high measurement error) variants
  \item We demonstrate that the mutation generating process (mutation profile) can be reliably retrieved from simulated NGS data.
  \item We also demonstrate that on the limited real data available, which consists of validated high confidence mutations, the results are equivalent
  \item What is needed is validation experiments that also allow for the estimation of true and false negative rates, in order to fully illuminate the utility of \batcave
  \item We have devoloped this algorithm as a post-calling evaluation tool to be used at the end of a variant calling pipeline
  \
  \item We believe the improved calibration achieved with this algorithm will encourage more thoughtful consideration of thresholds in various variant calling contexts
  \item The field believes that little information can be gained by wes rather than panels, but they are assuming the only information is mutations \cite{Hyman2015}
\end{itemize}

>>>>

\batcave is a post-call variant evalutation tool that uses MuTect output and the tumor-specific mutation profile to sharpen the posterior probability assigned to SNVs.
We demonstrated here that \batcave improves classification accuracy in simulated data, with the magnitude of the improvement dependent on the sequencing depth and the concentration of the mutation profile.

We report only precision-recall comparisons for real tumor data because in both cases the "gold" standard data sets consist of variants that were "called" by one or more variant detection pipelines, and confirmed to be present or absent.
These mutations are in some sense \textit{easy} to call, since at least one variant caller did in fact call them.
At the same time, there is no way to compute true and false negative fractions for these data sets.
As such, it is not surprising that the performance of MuTect and \batcave are similar because the universe of variants under consideration in these experiments are by definition "callable" by at least one variant detection pipeline, and MuTect is among the most sensitive and specific variant callers \cite{Griffith2015,Cibulskis2013}.
Deep sequencing experiments where a large random sample of uncalled potential variants, as well as a random sample of variant with no reads at all present, would give much-needed insight into the differences in statistical models in variant calling.

The \batcave algorithm focuses on the tri-nucleotide context and substitution type, which has been shown to have a large effect on site-specific mutation rates \cite{Martincorena2015,Hollstein2017}.
There are, however, many other types of \textit{genomic context} which can effect local mutation rates such as replication timing, expression level, and histone occupancy \cite{Buisson2019} .
These could be very useful in computing a local or regional mutation rate, as opposed to the genome-wide but tumor-specific mutation rate used by \batcave.

@@

\batcave approaches the set of variant calls returned from MuTect as two distinct sets of data.
The first is the set of mutations that pass all heuristics filters, and  have a very large posterior probability of being present given the observed read data.
These high-probability variants are generated from the mutation profile of the tumor, and measured with very small error.
This data allows us to reliably estimate the distribution of substitution types.
Variants with low probability given the read data are noisy measurements of samples from the same data generating distribution.
We use the estimated distribution of substitution types to adjust the prior probability of mutation for these lower probability variants.

@@
